---
title: "Data preparation"
Author: "Gabriel Ortega-Solís"
format: html
---
## Overview
This script collects and prepares all the data required for the analysis of the geometric attributes of protected areas. It includes loading necessary libraries, reading and processing various datasets (climate, landcover, NDVI, elevation), calculating additional variables, and merging all datasets into a final dataframe ready for analysis.

#### Libraries
```{r}
pacman::p_load(tidyverse, vegan, sf, GeoThinneR, parallel)
```

### Unzip data files

```{python}
!rm -rf /tmp/data
!mkdir -p /tmp/data
!rm -rf /tmp/continents
!mkdir -p /tmp/continents
!rm -rf /tmp/ecoregions
!mkdir -p /tmp/ecoregions
!unzip Data/data.zip -d /tmp/data/
!ls -l /tmp/data | wc -l
!unzip Data/continents.zip -d /tmp/continents
!unzip Data/ecoregions.zip -d /tmp/ecoregions
```

### Join Geopackage layers from previous scripts
```{r}
pacman::p_load(sf, tidyverse)
layer_list <- st_layers("edge_bck/wdpa_wdpaid_10000.gpkg")$name
layer_list

wdpaids <- read.csv("edge_bck/wdpaid.csv",
    header = FALSE
)[[1]]

lapply(layer_list, function(layer) {
    result_df <- lapply(wdpaids, function(wdpaid) {
        try({
            df <- st_read(
                str_glue("edge_bck/wdpa_wdpaid_{wdpaid}.gpkg"),
                layer = layer
            )
        })
        try({
            df <- st_drop_geometry(df)
        })
        return(df)
    }) %>%
        # Keep only data.frames and remove NULL/error objects
        .[sapply(., is.data.frame)] %>%
        bind_rows()

    # Write CSV file for this layer
    write_csv(result_df, str_glue("Data/output_{layer}.csv"))

    return(result_df)
})
```

### Import datasets
```{r}
cores <- 20
climate <- list.files("/tmp/data/", full.names = T, pattern = "climate_analysis") %>%
    mclapply(., read_csv, mc.cores = cores) %>%
    data.table::rbindlist(fill = T) %>%
    select(-1, -.geo) %>%
    filter(is.na(climate_histogram)) %>%
    select(-WDPAID, -climate_histogram) %>%
    rename(WDPAID = wdpaid) %>%
    mutate(WDPAID = as.factor(WDPAID)) %>%
    group_by(WDPAID) %>%
    summarise(across(everything(), mean, na.rm = TRUE))

landcover <- list.files("/tmp/data/", full.names = T, pattern = "landcover_stats") %>%
    mclapply(., read_csv, mc.cores = cores) %>%
    data.table::rbindlist(fill = T) %>%
    select(-1, -.geo) %>%
    mutate(WDPAID = as.factor(wdpaid)) %>%
    select(-wdpaid) %>%
    group_by(WDPAID) %>%
    summarise(across(everything(), mean, na.rm = TRUE))

ndvi <- list.files("/tmp/data/", full.names = T, pattern = "mean_sd_NDVI") %>%
    mclapply(., read_csv, mc.cores = cores) %>%
    data.table::rbindlist(fill = T) %>%
    select(-1, -.geo) %>%
    mutate(WDPAID = as.factor(WDPAID))

skip_cols <- c("id_1", "fixed")
data <- read_csv("Data/output_geometric_attributes.csv") %>%
    select(-all_of(skip_cols)) %>%
    mutate(ns_elongation = ns_length / ew_length) %>%
    rename(A = area) %>%
    left_join(read_csv("Data/output_buffer_neg_100m.csv") %>% select(-all_of(skip_cols)), by = "wdpaid") %>%
    rename(neg100 = area) %>%
    left_join(., read_csv("Data/output_buffer_neg_500m.csv") %>% select(-all_of(skip_cols)), by = "wdpaid") %>%
    rename(neg500 = area) %>%
    left_join(., read_csv("Data/output_buffer_neg_1000m.csv") %>% select(-all_of(skip_cols)), by = "wdpaid") %>%
    rename(neg1000 = area) %>%
    left_join(., read_csv("Data/output_buffer_neg_2000m.csv") %>% select(-all_of(skip_cols)), by = "wdpaid") %>%
    left_join(., read_csv("Data/output_elevation.csv") %>% select(-all_of(skip_cols)), by = "wdpaid") %>%
    rename(neg2000 = area) %>%
    rename(WDPAID = wdpaid, area = A) %>%
    mutate(WDPAID = as.factor(WDPAID)) %>%
    mutate(
        edge100m = area - neg100,
        edge500m = area - neg500,
        edge1000m = area - neg1000,
        edge2000m = area - neg2000,
        elevation_range = elevation_max - elevation_min,
        elevation_cv = elevation_sd / elevation_mean
    ) %>%
    mutate(edge_prop = edge1000m / area)

nrow(data)
length(unique(data$WDPAID))
names(data)
sapply(data, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE)
```

#### Climate

```{r}
# Process the climate data
cols <- paste0("cl", as.character(c(1:30)))

if (any(str_detect(colnames(climate), "^cl\\d+$"))) {
    cat("Column names already formatted.\n")
} else {
    colnames(climate) <- colnames(climate) %>% mgsub::mgsub(., pattern = as.character(c(1:30)), replacement = cols)
}

climate$shannon_climates <- apply(climate[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "shannon")
})

climate$gini_simpson_climates <- apply(climate[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "simpson")
})

climate$inv_simpson_climates <- apply(climate[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "invsimpson")
})

climate$n_climates <- vegan::specnumber(climate[, cols])

climate$Pielou_climates <- climate$shannon_climates / log(climate$n_climates)

climate$simpson_evenness_climate <- climate$inv_simpson_climates / climate$n_climates

length(unique(climate$WDPAID)) == nrow(climate)
sapply(climate, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE)
```

#### Landcover

```{r}
# Process the landcover data
cols <- paste0("lc", c("20", "30", "40", "60", "70", "80", "90", "100", "111", "112", "113", "114", "115", "116", "121", "122", "123", "124", "125", "126"))

if (any(str_detect(colnames(landcover), "^lc\\d+$"))) {
    cat("Column names already formatted.\n")
} else {
    colnames(landcover) <- colnames(landcover) %>% mgsub::mgsub(., pattern = c("20", "30", "40", "60", "70", "80", "90", "100", "111", "112", "113", "114", "115", "116", "121", "122", "123", "124", "125", "126"), replacement = cols)
}

landcover$shannon_landcover <- apply(landcover[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "shannon")
})

landcover$gini_simpson_landcover <- apply(landcover[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "simpson")
})

landcover$inv_simpson_landcover <- apply(landcover[, cols], 1, function(x) {
    x <- x[x > 0] # Remove zero values
    vegan::diversity(x, index = "invsimpson")
})

landcover$n_landcover <- vegan::specnumber(landcover[, cols])

landcover$Pielou_landcover <- landcover$shannon_landcover / log(landcover$n_landcover)

landcover$simpson_evenness_landcover <- landcover$inv_simpson_landcover / landcover$n_landcover

length(unique(landcover$WDPAID)) == nrow(landcover)
sapply(landcover, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE)
```

#### Built space
```{r}
landcover$built_space <- landcover$`50` / landcover$total_pixels
```

#### Elevation

```{r}
# elevation <- elevation %>%
#     group_by(WDPAID) %>%
#     summarise(across(everything(), mean)) %>%
#     mutate(
#         elevation_range = elevation_max - elevation_min,
#         elevation_cv = elevation_stdDev / elevation_mean
#     )
# length(unique(elevation$WDPAID)) == nrow(elevation)
```

#### NDVI - EVI

```{r}
ndvi <- ndvi %>%
    select(-year) %>%
    group_by(WDPAID) %>%
    summarise(
        EVI_p90_mean = mean(EVI_p90_mean),
        NDVI_p90_mean = mean(NDVI_p90_mean),
        EVI_p90_stdDev = sqrt(mean(EVI_p90_stdDev^2)),
        NDVI_p90_stdDev = sqrt(mean(NDVI_p90_stdDev^2))
    ) %>%
    mutate(
        cv_evi = EVI_p90_stdDev / EVI_p90_mean,
        cv_ndvi = NDVI_p90_stdDev / NDVI_p90_mean
    )

length(unique(ndvi$WDPAID)) == nrow(ndvi)
nrow(ndvi)
sapply(ndvi, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE)
```

#### Add continents and biomes
```{r}
# Make coordinates object
coords <- select(data, WDPAID, decimallongitude, decimallatitude) %>%
    mutate(lon = decimallongitude, lat = decimallatitude) %>%
    mutate(x = decimallongitude, y = decimallatitude) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
    st_make_valid() %>%
    unique()

# Load continent and biome data
continents <- st_read("/tmp/continents/continents.gdb")[, c("CONTINENT", "SHAPE")] %>%
    rename(continent = CONTINENT, geometry = SHAPE) %>%
    st_make_valid() %>%
    st_transform(., st_crs(coords))

plot(st_geometry(continents))
points(st_geometry(coords), col = "red", pch = 20, cex = 0.5)

biomes <- st_read("/tmp/ecoregions//wwf_ecoregions//wwf_terr_ecos.shp")[, c("REALM", "geometry")] %>%
    rename(realm = REALM) %>%
    st_make_valid() %>%
    st_transform(., st_crs(coords))

plot(st_geometry(biomes))
points(st_geometry(coords), col = "red", pch = 20, cex = 0.5)

# Use st_join with prepared spatial indexes for faster joins
coords <- coords %>% st_join(continents, join = st_intersects)
coords <- coords %>% st_join(biomes, join = st_intersects)

length(unique(coords$WDPAID)) == nrow(coords)
```


#### Latitude and longitude to cartesian coordinates

```{r}
# Apply lat-long transformation into cartesian coordinates
geo2cartesian <- function(data, lat, lon) {
    xyz <- lon_lat_to_cartesian(lat = data[[lat]], lon = data[[lon]])
    data[[lon]] <- xyz[, 1] # First column is x
    data[[lat]] <- xyz[, 2] # Second column is y
    data$Z <- xyz[, 3] # Third column is z
    return(data)
}

coords <- geo2cartesian(coords, lat = "y", lon = "x")

length(unique(coords$WDPAID)) == nrow(coords)
```

### Merge all datasets

```{r}
# Perform joins, handle missing values and calculate additional variables
final_data <- data %>%
    left_join(climate, by = "WDPAID") %>%
    left_join(landcover, by = "WDPAID") %>%
    left_join(ndvi, by = "WDPAID") %>%
    left_join(select(coords, -decimallatitude, -decimallongitude), by = "WDPAID") %>%
    mutate(
        ns_elongation = ns_length / ew_length,
        WDPAID = as.character(WDPAID)
    ) %>%
    select(-fixed, -Pielou_landcover, -matches("^lc\\d"), -matches("^cl\\d"))

# Check NAs per column
na_counts <- sapply(final_data, function(x) sum(is.na(x))) %>% sort(decreasing = TRUE)
print(na_counts)

# Find duplicated WDPAIDs
length(final_data$WDPAID[duplicated(final_data$WDPAID)])

saveRDS(final_data, "Data/wdpa.rds")

nrow(na.omit(final_data))
```
