---
title: "Protected Areas Geometric Attributes and Biodiversity"
author: 
    - name: Gabriel Ortega
    - name: Petr Keil
    - name: Michela Perrone
    - name: Daniela Mellado
    - name: Manuele Bazzichetto
abstract: |
    Protected areas (PAs) design emphasizes shapes closer to a circle to reduce border effects. Kunin (1997) shows that plots with elongated shapes have more biodiversity. Kunin's findings are based on the principle that closer things are more similar, so elongated plots (ergo elongated PAs) should include things that are more distant between them, which are more likely to be different. While we cannot record species richness (or another diversity metric) worldwide simultaneously and with spatial accuracy, metrics derived from Earth Observation have been proposed as alternative proxies for biodiversity monitoring. The Spectral Variation Hypothesis proposes that geographic areas with high heterogeneity in remote images present high environmental heterogeneity, and therefore, support more biodiversity. 
format: 
    html: 
        toc: true
        toc-depth: 2
        number-sections: true
        code-fold: true
---

# Introduction

-Protected areas purposes

-Spatial design of protected areas Protected areas (PAs) design emphasizes shapes closer to a circle to reduce border effects.

-Distance decay principle Kunin (1997) shows that plots with elongated shapes have more biodiversity. Kunin's findings are based on the principle that closer things are more similar, so elongated plots (ergo elongated PAs) should include things that are more distant between them, which are more likely to be different. While we cannot record species richness (or another diversity metric) worldwide simultaneously and with spatial accuracy, metrics derived from Earth Observation have been proposed as alternative proxies for biodiversity monitoring. The Spectral Variation Hypothesis proposes that geographic areas with high heterogeneity in remote images present high environmental heterogeneity, and therefore, support more biodiversity.

-Relationship between geometric attributes and habitat heterogeneity The Spectral Variation Hypothesis proposes that geographic areas with high heterogeneity in remote images present high environmental heterogeneity, and therefore, support more biodiversity.

-Here we address the relationship between geometric attributes and habitat heterogeneity in protected areas worldwide.

-The fundamental principle behind all our hypotheses is the distance decay of similarity (Tobler’s law, Tobler XXX, Nekola & White 2001), which states that closer locations are more similar in environmental conditions and species composition than distant areas.

# Hypotheses

1.  **Heterogeneity increases with area**: Larger protected areas encompass a broader range of habitats by including more distant locations, thereby increasing habitat heterogeneity due to the principle of distance decay of similarity.
2.  **Elongated PAs have more habitat heterogeneity**: Given constant area, elongated PAs cover longer environmental gradients than round PAs, increasing habitat heterogeneity due to the distance decay of similarity.
3.  **Fragmented PAs harbor more heterogeneity**: Given constant area, fragmented PAs (measured as the number of polygons) are expected to have higher heterogeneity (Fahrig, Chase).
4.  **Shape complexity increases heterogeneity**: Given constant area, PAs with higher shape complexity (measured as an area-to-perimeter ratio) are expected to have more heterogeneity.
5.  **North-South extent increases heterogeneity**: Habitat heterogeneity increases with increasing North-South extent, as PAs capture the global latitudinal gradient of temperature and precipitation. This effect is detectable above a threshold value of extent.
6.  **Wide elevational range**: Wider elevational ranges cover a higher variability in environmental conditions, increasing habitat heterogeneity.
7.  **Bioclimatic zonation**: Bigger areas include more diverse climates and thus more habitat heterogeneity.

We expect some of these metrics/hypotheses to be correlated with each other, and it may not be possible to separate their effects due to fundamental dependencies (Sizling et al. 2024).

# Methods

## Step 1: Protected areas dataset

We used the December 2024 version of the World Database of Protected Areas. We restricted the analyses to terrestrial protected areas with area higher than zero and whose status is one of the following: "Designated", "Inscribed", "Adopted" or "Established". Each PA is identified with a WDPAID assigned by the UN Environment Programme World Conservation Monitoring Centre. However, protected areas could receive multiple administrative designations according to different institutions, resulting in multiple WDPAID for overlapping areas. For instance, part of a national park could be also a natural monument and be listed twice in the database. To avoid problems with overlapping areas we dissolved PAs sharing the same WDPAID and then identified all of the intersecting areas.

```{sql}
-- 1. Create initial geometry aggregation with spatial index
CREATE MATERIALIZED VIEW admin.terrestrial_geometry_1 AS
SELECT
    wdpaid,
    ST_UNION(geometry) AS geometry
FROM public.wdpa_poly_dec2024
WHERE marine = '0'
  AND gis_area > 0
  AND status IN ('Designated', 'Inscribed', 'Adopted', 'Established')
GROUP BY wdpaid;
```

Add spatial index to speed up further processing.

```{sql}
CREATE INDEX tg1_geometry_idx ON admin.terrestrial_geometry_1 USING GIST(geometry);
```

Add an additional index on the unique identifier.

```{sql}
CREATE INDEX tg1_wdpaid_idx ON admin.terrestrial_geometry_1 (wdpaid);
```

Since some areas overlap, we created an additional table to identify the set of intersecting WDPAIDs per polygon.

```{sql}
-- 2. Create intersection analysis with optimized joins
CREATE MATERIALIZED VIEW admin.terrestrial_intersections_2 AS
SELECT
    a.wdpaid,
    ARRAY_AGG(DISTINCT b.wdpaid) AS intersecting_ids
FROM admin.terrestrial_geometry_1 a
JOIN admin.terrestrial_geometry_1 b 
  ON a.wdpaid <> b.wdpaid
  AND ST_INTERSECTS(a.geometry, b.geometry)
GROUP BY a.wdpaid;
```

Add a spatial index to the new table.

```{sql}
CREATE INDEX ti2_wdpaid_idx ON admin.terrestrial_intersections_2 (wdpaid);
```

We create a combined view of the two previous tables.

```{sql}
-- 3. Create final combined view
CREATE OR REPLACE VIEW admin.wdpaid_terrestrial_union AS
SELECT
    g.wdpaid,
    g.geometry,
    COALESCE(i.intersecting_ids, ARRAY[]::bigint[]) AS intersecting_ids
FROM admin.terrestrial_geometry_1 g
LEFT JOIN admin.terrestrial_intersections_2 i
  ON g.wdpaid = i.wdpaid;
```

With the previous view in place we can create a materialized view to pull data from it.

```{sql}
-- 4. Create public-facing materialized view
CREATE MATERIALIZED VIEW public.wdpaid_terrestrial_union AS
SELECT * FROM admin.wdpaid_terrestrial_union;
```

Add a spatial index to the materialized view.

```{sql}
-- Add final indexes
CREATE INDEX wtu_geometry_idx ON public.wdpaid_terrestrial_union USING GIST(geometry);
```

Add an index to the unique identifier in the materialized view.

```{sql}
CREATE INDEX wtu_wdpaid_idx ON public.wdpaid_terrestrial_union (wdpaid);
```

## Step 2: Characterization of Geometrical Dimensions

We first characterized the fundamental geometrical dimensions (axes) of the PAs. This step is essential to address potential collinearities among geometrical characteristics. For instance, elongated areas may be large, round areas may be small, and fragmented areas may have low area-perimeter ratios.

```{sql}
-- 1. Geometry metrics view and materialized view
CREATE OR REPLACE VIEW admin.matview_geometric_metrics
 AS
 SELECT data.wdpaid,
    st_area(data.geometry::geography) / 1000000.0::double precision AS area_km2,
        CASE
            WHEN mbr.short > 0::double precision THEN mbr.long / mbr.short
            ELSE 0::double precision
        END AS elongationratio,
    4.0::double precision * pi() * (st_area(data.geometry::geography) / 1000000.0::double precision) / NULLIF(power(st_perimeter(data.geometry::geography) / 1000.0::double precision, 2::double precision), 0::double precision) AS normcircularity,
    st_numgeometries(data.geometry) AS numpolygons,
    st_nrings(data.geometry) - st_numgeometries(data.geometry) AS numholes,
        CASE
            WHEN mbr.hull_area > 0::double precision THEN st_area(data.geometry::geography) / 1000000.0::double precision / mbr.hull_area
            ELSE 0::double precision
        END AS compactnessratio,
        CASE
            WHEN (st_perimeter(data.geometry::geography) / 1000.0::double precision) > 0::double precision AND (st_area(data.geometry::geography) / 1000000.0::double precision) > 0::double precision THEN 2.0::double precision * ln(st_perimeter(data.geometry::geography) / 1000.0::double precision) / ln(st_area(data.geometry::geography) / 1000000.0::double precision)
            ELSE NULL::double precision
        END AS fractaldimension
   FROM wdpaid_terrestrial_union data
     LEFT JOIN LATERAL ( SELECT st_area(st_convexhull(data.geometry)::geography) / 1000000.0::double precision AS hull_area,
            st_distance(p1.p1::geography, p2.p2::geography) / 1000.0::double precision AS long,
            st_distance(p2.p2::geography, p3.p3::geography) / 1000.0::double precision AS short
           FROM st_orientedenvelope(data.geometry) oe(oe),
            LATERAL st_pointn(st_exteriorring(oe.oe), 1) p1(p1),
            LATERAL st_pointn(st_exteriorring(oe.oe), 2) p2(p2),
            LATERAL st_pointn(st_exteriorring(oe.oe), 3) p3(p3)) mbr ON true;

CREATE MATERIALIZED VIEW IF NOT EXISTS public.mv_geometric_metrics
 SELECT * FROM admin.matview_geometric_metrics;

----------------------------------------------------------------------------------------------------------
-- 2. Extension metrics view and materialized view
CREATE OR REPLACE VIEW admin.matview_spatial_extension
 AS
 SELECT wdpaid_terrestrial_union.wdpaid,
    st_distance(st_makepoint(st_xmin(box.box::box3d), (st_ymin(box.box::box3d) + st_ymax(box.box::box3d)) / 2::double precision)::geography, st_makepoint(st_xmax(box.box::box3d), (st_ymin(box.box::box3d) + st_ymax(box.box::box3d)) / 2::double precision)::geography) / 1000::double precision AS ew_length,
    st_distance(st_makepoint((st_xmin(box.box::box3d) + st_xmax(box.box::box3d)) / 2::double precision, st_ymin(box.box::box3d))::geography, st_makepoint((st_xmin(box.box::box3d) + st_xmax(box.box::box3d)) / 2::double precision, st_ymax(box.box::box3d))::geography) / 1000::double precision AS ns_length,
    (( SELECT max(st_distance(a.geom::geography, b.geom::geography)) AS max
           FROM st_dumppoints(st_convexhull(wdpaid_terrestrial_union.geometry)) a(path, geom),
            st_dumppoints(st_convexhull(wdpaid_terrestrial_union.geometry)) b(path, geom))) / 1000::double precision AS maxlength_km
   FROM wdpaid_terrestrial_union
     CROSS JOIN LATERAL st_envelope(wdpaid_terrestrial_union.geometry) box(box);

CREATE MATERIALIZED VIEW IF NOT EXISTS public.mv_extension_metrics
 SELECT * FROM admin.matview_spatial_extension; 

-----------------------------------------------------------------------------------------------------------
-- 3. Combined view
CREATE OR REPLACE VIEW public."MOBI_WDPA_land_combined_metrics"
 AS
 SELECT base.wdpaid,
    base.intersecting_ids,
    geom.area_km2 AS area,
    geom.elongationratio,
    geom.normcircularity,
    geom.numpolygons,
    geom.numholes,
    geom.compactnessratio,
    geom.fractaldimension,
    ext.ew_length,
    ext.ns_length,
    ext.maxlength_km AS maxlength
   FROM wdpaid_terrestrial_union base
     JOIN mv_geometric_metrics geom USING (wdpaid)
     JOIN mv_extension_metrics ext USING (wdpaid);
```

We extracted elevation data from Copernicus WorldEM-30.

```{javascript}
// Author: Gabriel Ortega

// This script is organized in four sections:
// 1. SETTINGS: Contains information about the input data and some pre-processing steps.
// 2. FUNCTION(S): Main function(s) to apply.
// 3. EXECUTION: Apply the function(s) to the input data.
// 4. EXPORT RESULTS: Export outside of EarthEngine.

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 1. SETTINGS

// Polygons collection
var pols = ee.FeatureCollection("WCMC/WDPA/current/polygons"); // Explicitly define WDPA polygons
// Unique polygons identifier
var uniqueID = 'WDPAID';

// Elevation Dataset
var elevationDataset = ee.ImageCollection("COPERNICUS/DEM/GLO30"); // Use the Copernicus DEM GLO 30m
var elevationBand = 'DEM'; // Band name for elevation is 'DEM'
var waterMaskBand = 'WBM'; // Band name for water mask is 'WBM'

// Create a bounding box if necessary

// Original bounding box coordinates
var xmin = -180;    // left
var ymin = 90;   // top
var xmax = 180;  // right
var ymax = -90;    // bottom

// Define which box you want
// Select box (0 = full, 1-16 = subdivisions)
var whichBox = 16;  // Change this value

// Initialize bbox variable
var bbox;

if (whichBox === 0) {
  // Full bounding box
  bbox = ee.Geometry.BBox(xmin, ymin, xmax, ymax);
} else {
  // Grid configuration (4x4)
  var xSegments = 4;
  var ySegments = 4;
  
  // Calculate step sizes
  var xStep = (xmax - xmin)/xSegments;  // 45° per column
  var yStep = (ymax - ymin)/ySegments;  // -22.5° per row
  
  // Convert whichBox (1-16) to grid indices
  var adjustedIndex = whichBox - 1;
  var row = Math.floor(adjustedIndex / xSegments);  // 0-3 (bottom to top)
  var col = adjustedIndex % xSegments;              // 0-3 (left to right)
  
  // Reverse row order (original grid is top to bottom)
  var yIndex = (ySegments - 1) - row;
  
  // Calculate coordinates
  var xStart = xmin + col * xStep;
  var xEnd = xStart + xStep;
  var yStart = ymin + yIndex * yStep;
  var yEnd = yStart + yStep;
  
  bbox = ee.Geometry.BBox(xStart, yStart, xEnd, yEnd);
}


// Name of output file
var fileName = 'elevation_stats_WDPA_'+xmin+'_'+ymin+'_'+xmax+'_'+ymax+'_bbox'+whichBox;

// Filter polygons dataset as required
var filteredPols = pols
  .filterBounds(bbox)
  .filter(
    ee.Filter.and(
      ee.Filter.notNull([uniqueID]),
      ee.Filter.eq('MARINE', '0'),
      ee.Filter.gt('GIS_AREA', 0),  // Added gis_area > 0 filter
      ee.Filter.inList('STATUS', [  // Added status filter
        'Designated',
        'Inscribed',
        'Adopted',
        'Established'
      ])
    )
  );


// Group polygons by WDPA_PID and combine their geometries.
var uniquePolID = filteredPols.aggregate_array(uniqueID).distinct().map(function(id) {
  var collection = filteredPols.filter(ee.Filter.eq(uniqueID, id));
  // Combine geometries using a unary union (merging all geometries in the collection).
  var combinedGeometry = collection.geometry().dissolve(); // Use dissolve to combine all geometries
  return ee.Feature(combinedGeometry, {WDPAID: id});
});

// Convert to FeatureCollection
var uniquePolygons = ee.FeatureCollection(uniquePolID);

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 2. FUNCTION(S)

var processElevation = function() {

  var features = uniquePolygons.map(function(feature) {
    var combinedGeometry = feature.geometry();

    // Get the elevation and water mask bands.  Since elevationDataset is a collection, we need to mosaic it.
    var elevation = elevationDataset.select(elevationBand).mosaic();
    var waterMask = elevationDataset.select(waterMaskBand).mosaic();

    // Apply the water mask:
    // WBM values: 0=No water, 1=Ocean, 2=Lake, 3=River.
    // We want to include land, which is WBM == 0.
    var maskedElevation = elevation.updateMask(waterMask.eq(0)); // Keep only where WBM == 0

    // Calculate elevation statistics.
    var stats = maskedElevation.reduceRegion({
      reducer: ee.Reducer.min()
        .combine(ee.Reducer.max(), null, true)
        .combine(ee.Reducer.mean(), null, true)
        .combine(ee.Reducer.stdDev(), null, true),
      geometry: combinedGeometry,
      scale: 30,
      maxPixels: 1e13,
      bestEffort: true
    });

    var renamedStats = {
        elevation_min: stats.get('DEM_min'),
        elevation_max: stats.get('DEM_max'),
        elevation_mean: stats.get('DEM_mean'),
        elevation_stdDev: stats.get('DEM_stdDev'),
    };
    return ee.Feature(null, renamedStats).set(uniqueID, feature.get(uniqueID));
  });

  return ee.FeatureCollection(features);
};

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 3. EXECUTION

var allFeatures = processElevation();

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 4. EXPORT RESULTS

Export.table.toDrive({
  collection: allFeatures,
  description: fileName,
  fileFormat: 'CSV',
  folder: 'Earthengine_exports'
});
```

We extracted climate data from ...

```{javascript}
// Authors: Gabriel Ortega

// This script is organized in four sections:
// 1. SETTINGS: Contains information about the input data and some pre-processing steps.
// 2. FUNCTION(S): Main function(s) to apply.
// 3. EXECUTION: Apply the function(s) to the input data.
// 4. EXPORT RESULTS: Export outside of EarthEngine.

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 1. SETTINGS

// Polygons collection
var pols = ee.FeatureCollection("WCMC/WDPA/current/polygons"); // Explicitly define WDPA polygons
// Unique polygons identifier
var uniqueID = 'WDPAID';
// Climate Dataset
var climateDataset = ee.Image('projects/gortega-research/assets/koppen_geiger_0p01'); // Koppen-Geiger climate zones
var climateBand = 'b1'; // Band name for climate classification

// Create a bounding box if necessary

// Original bounding box coordinates
var xmin = -180;    // left
var ymin = 90;   // top
var xmax = 180;  // right
var ymax = -90;    // bottom

// Define which box you want
// Select box (0 = full, 1-16 = subdivisions)
var whichBox = 16;  // Change this value

// Initialize bbox variable
var bbox;

if (whichBox === 0) {
  // Full bounding box
  bbox = ee.Geometry.BBox(xmin, ymin, xmax, ymax);
} else {
  // Grid configuration (4x4)
  var xSegments = 4;
  var ySegments = 4;
  
  // Calculate step sizes
  var xStep = (xmax - xmin)/xSegments;  // 45° per column
  var yStep = (ymax - ymin)/ySegments;  // -22.5° per row
  
  // Convert whichBox (1-16) to grid indices
  var adjustedIndex = whichBox - 1;
  var row = Math.floor(adjustedIndex / xSegments);  // 0-3 (bottom to top)
  var col = adjustedIndex % xSegments;              // 0-3 (left to right)
  
  // Reverse row order (original grid is top to bottom)
  var yIndex = (ySegments - 1) - row;
  
  // Calculate coordinates
  var xStart = xmin + col * xStep;
  var xEnd = xStart + xStep;
  var yStart = ymin + yIndex * yStep;
  var yEnd = yStart + yStep;
  
  bbox = ee.Geometry.BBox(xStart, yStart, xEnd, yEnd);
}


// Name of output file
var fileName = 'climate_analysis_WDPA_'+xmin+'_'+ymin+'_'+xmax+'_'+ymax+'_bbox'+whichBox;

// Filter polygons dataset as required
var filteredPols = pols
  .filterBounds(bbox)
  .filter(
    ee.Filter.and(
      ee.Filter.notNull([uniqueID]),
      ee.Filter.eq('MARINE', '0'),  // Keep only land areas
      ee.Filter.gt('GIS_AREA', 0),  // Keep gis_area > 0 filter
      ee.Filter.inList('STATUS', [  // Keep status in list
        'Designated',
        'Inscribed',
        'Adopted',
        'Established'
      ])
    )
  );


// Group polygons by WDPAID and combine their geometries.
var groupedPolygons = filteredPols.aggregate_array(uniqueID).distinct().map(function(id) {
  var collection = filteredPols.filter(ee.Filter.eq(uniqueID, id));
  // Combine geometries using a unary union (merging all geometries in the collection).
  var combinedGeometry = collection.geometry().dissolve(); // Use dissolve to combine all geometries
  return ee.Feature(combinedGeometry, {WDPAID: id});
});

// Convert to FeatureCollection
var uniquePolygons = ee.FeatureCollection(groupedPolygons);

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 2. FUNCTION(S)
var processData = function() {

  var features = uniquePolygons.map(function(feature) {
    var combinedGeometry = feature.geometry();

    // Get the climate data.
    var climate = climateDataset.select(climateBand);

    // Use a histogram to get the pixel counts for each climate category within the combined polygon.
    var climateStats = climate.reduceRegion({
        reducer: ee.Reducer.histogram(),
        geometry: combinedGeometry,
        scale: 100,
        maxPixels: 1e13
    });

    // Get the total number of pixels in the combined polygon.
    var totalPixels = climate.reduceRegion({
        reducer: ee.Reducer.count(),
        geometry: combinedGeometry,
        scale: 100,
        maxPixels: 1e13
    }).get(climateBand);

    var results = {
        climate_histogram: climateStats.get(climateBand),
        total_pixels: totalPixels
    };

    return ee.Feature(null, results).set(uniqueID, feature.get(uniqueID));
  });

  return ee.FeatureCollection(features);
};

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 3. EXECUTION

var allFeatures = processData();

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 4. EXPORT RESULTS
Export.table.toDrive({
  collection: allFeatures,
  description: fileName,
  fileFormat: 'CSV',
  folder: 'Earthengine_exports'
});
```

Finally we extracted NDVI data from

```{javascript}
// Authors: Gabriel Ortega & Michela Perrone

// This script is organized in four sections:
// 1. SETTINGS: Contains information about the input data and some pre-processing steps.
// 2. FUNCTION(S): Main function(s) to apply.
// 3. EXECUTION: Apply the function(s) to the input data.
// 4. EXPORT RESULTS: Export outside of EarthEngine.

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 1. SETTINGS

// Rasters collection
var collection = 'MODIS/061/MOD13Q1';

// Polygons collection
var pols = ee.FeatureCollection('WCMC/WDPA/current/polygons');

// Unique polygons identifier
var uniqueID = 'WDPAID';

// Set years of interest
var startYear = 2000;
var endYear = 2024;

// Select bands of interest
var bands = ['NDVI', 'EVI'];
var qualityBand = 'SummaryQA';
var qualityValue = 0;
var scaleFactor = 0.0001;

// Create a bounding box if necessary

// Original bounding box coordinates
var xmin = -180;    // left
var ymin = 90;   // top
var xmax = 180;  // right
var ymax = -90;    // bottom

// Define which box you want
// Select box (0 = full, 1-16 = subdivisions)
var whichBox = 16;  // Change this value

// Initialize bbox variable
var bbox;

if (whichBox === 0) {
  // Full bounding box
  bbox = ee.Geometry.BBox(xmin, ymin, xmax, ymax);
} else {
  // Grid configuration (4x4)
  var xSegments = 4;
  var ySegments = 4;
  
  // Calculate step sizes
  var xStep = (xmax - xmin)/xSegments;  // 45° per column
  var yStep = (ymax - ymin)/ySegments;  // -22.5° per row
  
  // Convert whichBox (1-16) to grid indices
  var adjustedIndex = whichBox - 1;
  var row = Math.floor(adjustedIndex / xSegments);  // 0-3 (bottom to top)
  var col = adjustedIndex % xSegments;              // 0-3 (left to right)
  
  // Reverse row order (original grid is top to bottom)
  var yIndex = (ySegments - 1) - row;
  
  // Calculate coordinates
  var xStart = xmin + col * xStep;
  var xEnd = xStart + xStep;
  var yStart = ymin + yIndex * yStep;
  var yEnd = yStart + yStep;
  
  bbox = ee.Geometry.BBox(xStart, yStart, xEnd, yEnd);
}


// Name of output file
var fileName = 'mean_sd_NDVI_EVI_WDPA_' + xmin + '_' + ymin + '_' + xmax + '_' + ymax + '_' + startYear + '_' + endYear+'_bbox'+whichBox;

// Filter polygons dataset as required
var filteredPols = pols
  .filterBounds(bbox)
  .filter(
    ee.Filter.and(
      ee.Filter.notNull([uniqueID]),
      ee.Filter.eq('MARINE', '0'),
      ee.Filter.gt('GIS_AREA', 0),  // Added gis_area > 0 filter
      ee.Filter.inList('STATUS', [  // Added status filter
        'Designated',
        'Inscribed',
        'Adopted',
        'Established'
      ])
    )
  );


// Group polygons by WDPA_PID and combine their geometries.
var uniquePolID = filteredPols.aggregate_array(uniqueID).distinct().map(function(id) {
  var collection = filteredPols.filter(ee.Filter.eq(uniqueID, id));
  // Combine geometries using a unary union (merging all geometries in the collection).
  var combinedGeometry = collection.geometry().dissolve(); // Use dissolve to combine all geometries
  return ee.Feature(combinedGeometry, {WDPAID: id});
});

// Convert to FeatureCollection
var uniquePolygons = ee.FeatureCollection(uniquePolID);

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 2. FUNCTION(S)

var processYear = function(year) {
  var startDate = ee.Date.fromYMD(year, 1, 1); // Indicate a starting date
  var endDate = ee.Date.fromYMD(ee.Number(year).add(1), 1, 1); // Indicate an ending date

  // 2.1. Image Processing:
  var processedImage = ee.ImageCollection(collection)
    .filterDate(startDate, endDate) // Filter start and end dates
    .select(bands.concat([qualityBand])) // Add the quality band to selection
    .map(function(image) {
      var qualityMask = image.select(qualityBand).eq(qualityValue); // Create a quality mask
      return image.updateMask(qualityMask).select(bands); // Apply the quality mask to each image
    })
    .reduce(ee.Reducer.percentile([90])) // Reduce all the overlapped image-pixels to a single one
    .multiply(scaleFactor); // Apply the corresponding scale factor
  
  // 2.2. Feature Generation:
  var features = uniquePolygons.map(function(feature) {
    var combinedGeometry = feature.geometry();
    
    // 2.3. Get variables of interest:
    var stats = processedImage.reduceRegion({
      reducer: ee.Reducer.mean().combine(ee.Reducer.stdDev(), null, true),
      geometry: combinedGeometry,
      scale: 250,        // Cell size
      maxPixels: 1e13,    // Increase if needed for large polygons
      bestEffort: true
    });

    // 2.4. Return annual results
    return ee.Feature(null, stats).set(uniqueID, feature.get(uniqueID)).set('year', year);
  });

  // 2.5. Return all results
  return ee.FeatureCollection(features);
};

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 3. EXECUTION

var allFeatures = ee.FeatureCollection(
  ee.List.sequence(startYear, endYear).map(processYear).flatten()
).flatten();

////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////////
// 4. EXPORT RESULTS

Export.table.toDrive({
  collection: allFeatures,
  description: fileName,
  fileFormat: 'CSV',
  folder: 'Earthengine_exports'
});
```

Merge data

```{bash}
mkdir -p /tmp/data
unzip data.zip -d /tmp/data/
```

### Data preparation

```{r}
pacman::p_load(tidyverse)
climate <- list.files("/tmp/data/", full.names = T, pattern = "climate_analysis") %>% lapply(., read_csv) %>% data.table::rbindlist(fill = T) %>% select(-1, -.geo)
elevation <- list.files("/tmp/data/", full.names = T, pattern = "elevation_stats") %>% lapply(., read_csv) %>% data.table::rbindlist(fill = T) %>% select(-1, -.geo)
ndvi <- list.files("/tmp/data/", full.names = T, pattern = "mean_sd_NDVI") %>% lapply(., read_csv) %>% data.table::rbindlist(fill = T) %>% select(-1, -.geo)
data <- list.files("/tmp/data/", full.names = T, pattern = "^MOBI_") %>% read_csv() %>% rename(WDPAID = wdpaid)
```

#### Climate

```{r}
# Process the climate data to extract and calculate relevant metrics
climate <- climate %>%
    mutate(
        # Extract the minimum bucket value from the climate histogram using regex
        bucketMin = as.numeric(str_extract(climate_histogram, "bucketMin=([0-9.]+)") %>% str_extract("[0-9.]+")),
        
        # Extract the bucket width from the climate histogram using regex
        bucketWidth = as.numeric(str_extract(climate_histogram, "bucketWidth=([0-9.]+)") %>% str_extract("[0-9.]+")),
        
        # Extract the histogram values as a numeric list using regex and string manipulation
        histogram = str_extract(climate_histogram, "(?<=histogram\\=)\\[.*?\\]") %>%
            str_remove_all("\\[|\\]") %>%
            str_split(",\\s*") %>%
            map(as.numeric),
        
        # Calculate the number of climate categories (bins) in the histogram
        n_climates = map_dbl(histogram, length),
        
        # Convert histogram values to numeric for further processing
        histogram_numeric = map(histogram, as.numeric)
    ) %>%
    group_by(WDPAID) %>%
    summarise(
        # Calculate the mean values for numeric columns (bucketMin, bucketWidth, n_climates)
        across(c(bucketMin, bucketWidth, n_climates), mean),
        
        # Compute the element-wise average of histogram lists across grouped data
        histogram_numeric = list(
            reduce(map(histogram_numeric, ~ {
                vec <- .x
                length(vec) <- max(lengths(histogram_numeric)) # Ensure all vectors have the same length
                replace_na(vec, 0) # Replace NA values with 0
            }), `+`) / n()
        )
    ) %>%
    mutate(
        # Calculate Shannon diversity index for each histogram
        shannon = map_dbl(histogram_numeric, ~ {
            counts <- .x[.x > 0] # Filter out zero counts
            if(length(counts) == 0) return(0) # Return 0 if no counts exist
            p <- counts / sum(counts) # Calculate proportions
            -sum(p * log(p)) # Compute Shannon index
        }),
        
        # Calculate Simpson diversity index for each histogram
        simpson = map_dbl(histogram_numeric, ~ {
            counts <- .x[.x > 0] # Filter out zero counts
            if(length(counts) == 0) return(0) # Return 0 if no counts exist
            p <- counts / sum(counts) # Calculate proportions
            1 - sum(p^2) # Compute Simpson index
        }),
        
        # Calculate evenness index for each histogram
        evenness = map_dbl(histogram_numeric, ~ {
            counts <- .x[.x > 0] # Filter out zero counts
            if(length(counts) == 0) return(1) # Return 1 if no counts exist
            p <- counts / sum(counts) # Calculate proportions
            H <- -sum(p * log(p)) # Compute Shannon index
            S <- length(counts) # Number of categories
            ifelse(S > 1, H / log(S), 1) # Compute evenness index
        })
    )
```

#### Elevation

```{r}
elevation <- elevation %>%
    group_by(WDPAID) %>%
    summarise(across(everything(), mean))
```

#### NDVI - EVI

```{r}
ndvi <- ndvi %>%
select(-year) %>%
group_by(WDPAID) %>%
summarise(EVI_p90_mean = mean(EVI_p90_mean),
NDVI_p90_mean = mean(NDVI_p90_mean),
EVI_p90_stdDev = sqrt(mean(EVI_p90_stdDev^2)),
NDVI_p90_stdDev = sqrt(mean(NDVI_p90_stdDev^2))) %>%
mutate(cv_evi = EVI_p90_stdDev/EVI_p90_mean,
cv_ndvi = NDVI_p90_stdDev/NDVI_p90_mean)
```

### Merge all datasets

```{r}
# Perform joins and handle missing values
data <- data %>%
  left_join(climate, by = "WDPAID") %>%
  left_join(elevation, by = "WDPAID") %>%
  left_join(ndvi, by = "WDPAID") %>%
  drop_na()
 # Use drop_na() to remove rows with missing values
```

## Step 2: Analysis

-   PCA with all attributes of protected areas.

```{r}
# Load required packages
pacman::p_load(ggplot2, FactoMineR, factoextra, dplyr)

# Prepare data for PCA
pca_data <- data %>% na.omit() %>%
  dplyr::select(-matches("WDPAID|^intersecting|^bucket|histogram_numeric|stdDev$|^EVI|^NDVI|cv_evi|cv_ndvi")) %>%
  filter(area > 10)

# Perform PCA
pca_result <- PCA(pca_data, scale.unit = TRUE, graph = FALSE)

# Visualize PCA results
fviz_pca_biplot(pca_result, 
                repel = TRUE, 
                col.var = "blue", 
                col.ind = "gray",
                title = "PCA Biplot of Geometric Attributes")
```

-   **H1**: Test the relationship between elongation and CV-EVI (and CV-NDVI) while controlling for area.
-   **H2**: Test the marginal effect of length in the North-South axis.
-   **H3**: \[Details to be added\].
-   **H4**: \[Details to be added\].

## Interpreting most important variables from random forest

-   Identify the threshold for North-South extent and CV-EVI.

```{r}
# Load required package
pacman::p_load(segmented)

# Fit linear model
lm_model <- lm(cv_evi ~ ns_length, data = data)

# Apply segmented regression
segmented_model <- segmented(lm_model, seg.Z = ~ns_length)

# Extract breakpoints
breakpoints <- segmented_model$psi[, "Est."]
print(breakpoints)

# Get fitted values
fitted_values <- fitted(segmented_model)
```

# Results

## Figure 1

PCA figure goes here.

```{r}
# Create density surface plot with variable importance arrows
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
    stat_density_2d(aes(fill = ..level..), geom = "polygon", color = NA, alpha = 0.7) +  # Density surface
    scale_fill_viridis_c() +
    geom_segment(data = pca_loadings, aes(x = 0, y = 0, xend = PC1 * 5, yend = PC2 * 5),
                 arrow = arrow(length = unit(0.2, "cm")), color = "red") +
    geom_text(data = pca_loadings, aes(x = PC1 * 5, y = PC2 * 5, label = rownames(pca_loadings)),
              hjust = 1.2, vjust = 1.2, color = "red") +
    labs(x = "Principal Component 1", y = "Principal Component 2") +
    ggpubr::theme_pubr()
```

## Figure XX

-Interpreting most important variables

```{r}
# Plot results
library(ggplot2)
plot_data <- data.frame(ns_length = data$ns_length, cv_evi = data$cv_evi, Fitted = fitted_values)

ggplot(plot_data, aes(x = ns_length, y = cv_evi)) +
    geom_point() + 
    geom_line(aes(y = Fitted), color = "blue") + 
    geom_vline(xintercept = breakpoints, linetype = "dashed", color = "red") + 
    labs(title = "Segmented Regression", x = "North-South Length", y = "CV-EVI") + 
    ggpubr::theme_pubr()
```

# Discussion

\[Discussion content to be added.\]

Nature protection often has complex goals: Šumava National Park and Czech Switzerland National Park are prime examples of protected areas where protecting landscape character is almost as important as protecting habitats and biodiversity.